Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-02-09 paper/zhang-cook-laa-langrene-menendez.Rmd
changes after talking to patricia

In projection pursuit the optimisation problem aims at finding the global and local maxima that give interesting projections according to an index function. That is, it starts with a given randomly selected basis $\mathbf{A}_1$ and aims at finding an optimal final projection basis $\mathbf{A}_T$ in each iteration that satisfies the following optimisation problem:

@@

\noindent where $f$, $\mathcal{A}$, and $\mathbf{X}$ are defined as in the previous section, and  $I_d$ is the $d$-dimensional identity matrix and the constraint requires the projection bases $\mathbf{A}$ to be orthonormal. Several features of this optimisation are worth-noticing. 1) The optimisation is constrained and the space of the bases is a $d$-dimensional sphere. 2) The objective function may not be differentiable for a constructed index function. 3) Although finding the global maximum is the goal of an optimisation problem, it is also interesting to inspect local maximum in projection pursuit since it could present unexpected interesting projections. 4) The optimisation procedure needs to be easy to compute since the tour animation needs to be played in real-time during the optimisation.

@@

Before making diagnostic plots, information generated by the optimisation need to be collected and a pre-defined data structure solidifies the information being collected and streamlines the subsequent plot-making.

@@

```{r toy-interp, fig.cap = "An inspection of the index values as the optimisation progress for two optimisers: SA (left) and SAJO (right) using trace plot.The holes index is optimised for a 2D projection problem on the six-variable dataset \\code{boa6}. The lines indicate the interpolation and the dots indicate new targets generated by the optimisers. The interpolation with both optimisers is smooth while the on the panel, SAJO is observed to first pass through some bases with higher index value before reach the target bases during Time 76 to 130."}
```

<!-- The right hand side has the jump out, which has a lower index value - also explain this in the text. -->

<!-- Both traces are smooth while the change of index value in each iteration may not be monotonic. -->

Another interesting diagnostic is to examine how the index value improves between interpolating bases since the projection on these bases is played by the tour animation. Trace plots are created with plotting the index value of the interpolation points against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisation routine as Figure \ref{fig:toy-search}. The trace plot is smooth and an interesting discovery is that in SAJO, the interpolation in iteration 6-8 is not monotonic. Sometimes, this non-monotonicity is acceptable since the target basis does not necessarily have a higher index value than the current one, while in some cases, the optimisation will be better off if these intermediate basis can be used as the start of the next iteration (see section \ref{monotonic} for a detailed discussion on this).

@@

```{r toy-pca-animated, fig.cap = "Six frames selected from the animation of search paths in the PCA-projected basis space. With animation, the start and end position of each optimisation routine is clearly identified. The full video of the animation can be found at \\url{https://vimeo.com/user132007430/review/504242845/b73f37175a}"}
```

@@

```{r toy-tour, fig.cap="Six frames selected from the tour view in the original 5D space for the same 1D projection problem as Figure \\ref{fig:toy-pca} and \\ref{fig:toy-pca-animated}. The full video of the animation can be found at \\url{https://vimeo.com/user132007430/review/504328122/9be84db563}"}
```

>>>>

In projection pursuit the optimisation problem aims at finding the global and local maxima that give interesting projections according to an index function. That is, it starts with a given randomly selected basis $\mathbf{A}_1$ and aims at finding an optimal final projection basis $\mathbf{A}_T$ that satisfies the following optimisation problem:

@@

\noindent where $f$, $\mathcal{A}$, and $\mathbf{X}$ are defined as in the previous section,  $I_d$ is the $d$-dimensional identity matrix and the constraint ensures the projection bases $\mathbf{A}$ to be orthonormal. It is worth noticing the following: 1) The optimisation is constrained and the basis space is a $d$-dimensional sphere. 2) There may be index functions for which the objective function might not be differentiable. 3) While finding the global optimum is the goal of the optimisation problem, interesting projections may also appear in the local optimum. 4) The optimisation should be computationally agile since the tour animation is viewed by the users during the optimisation.

@@

A data structure for diagnosing optimisers in projection pursuit guided tour is first defined. With a fixed data structure, the diagnostic plots presented later in this paper can be easily applied to diagnose other optimisation procedures.

@@

```{r toy-interp, fig.cap = "An inspection of the index values as the optimisation progress for two optimisers: SA (left) and SAJO (right) using trace plot.The holes index is optimised for a 2D projection problem on the six-variable dataset \\code{boa6}. The lines indicate the interpolation and the dots indicate new targets generated by the optimisers. The interpolation with both optimisers is smooth while the in the left panel, SAJO is observed to first pass through some bases with higher index value before reach the target bases during Time 76 to 130."}
```

Another interesting diagnostic is to examine how the index value improves between interpolating bases since the projection on these bases is played by the tour animation. Trace plots are created by plotting the index value against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisers as Figure \ref{fig:toy-search} and one can observe that the trace is smooth in both cases. It may seem bizarre at the first sight that the interpolation in the left panel first passes some bases with higher index value before decreases to a lower target. This happens because on the one hand, the probabilistic acceptance in SAJO implies that some worse bases will be accepted by the optimiser. On the other hand, guided tour interpolate between the current and target basis to provide a smooth transition between projections. These two facts indicate that a non-monotonic interpolation can't be avoided for SAJO. Later, in section \ref{monotonic}, there will be a discussion on improving the non-monotonic interpolation for SA.

@@

```{r toy-pca-animated, fig.cap = "Six frames selected from the animation of search paths in the PCA-projected basis space. With animation, the start and end position of each optimisation routine is clearly identified. The full video of the animation can be found at \\url{https://vimeo.com/user132007430/review/504242845/b73f37175a}."}
```

@@

```{r toy-tour, fig.cap="Six frames selected from the tour view in the original 5D space for the same 1D projection problem as Figure \\ref{fig:toy-pca} and \\ref{fig:toy-pca-animated}. The full video of the animation can be found at \\url{https://vimeo.com/user132007430/review/504328122/9be84db563}."}
```

### 2021-02-09 paper/zhang-cook-laa-langrene-menendez.tex
changes after talking to patricia

In projection pursuit the optimisation problem aims at finding the
global and local maxima that give interesting projections according to
an index function. That is, it starts with a given randomly selected
basis \(\mathbf{A}_1\) and aims at finding an optimal final projection
basis \(\mathbf{A}_T\) in each iteration that satisfies the following
optimisation problem:

@@

\noindent where \(f\), \(\mathcal{A}\), and \(\mathbf{X}\) are defined
as in the previous section, and \(I_d\) is the \(d\)-dimensional
identity matrix and the constraint requires the projection bases
\(\mathbf{A}\) to be orthonormal. Several features of this optimisation
are worth-noticing. 1) The optimisation is constrained and the space of
the bases is a \(d\)-dimensional sphere. 2) The objective function may
not be differentiable for a constructed index function. 3) Although
finding the global maximum is the goal of an optimisation problem, it is
also interesting to inspect local maximum in projection pursuit since it
could present unexpected interesting projections. 4) The optimisation
procedure needs to be easy to compute since the tour animation needs to
be played in real-time during the optimisation.

@@

\caption[An inspection of the index values as the optimisation progress for two optimisers]{An inspection of the index values as the optimisation progress for two optimisers: SA (left) and SAJO (right) using trace plot.The holes index is optimised for a 2D projection problem on the six-variable dataset \code{boa6}. The lines indicate the interpolation and the dots indicate new targets generated by the optimisers. The interpolation with both optimisers is smooth while the on the panel, SAJO is observed to first pass through some bases with higher index value before reach the target bases during Time 76 to 130.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

Another interesting diagnostic is to examine how the index value
improves between interpolating bases since the projection on these bases
is played by the tour animation. Trace plots are created with plotting
the index value of the interpolation points against time. Figure
\ref{fig:toy-interp} presents the trace plot of the same optimisation
routine as Figure \ref{fig:toy-search}. The trace plot is smooth and an
interesting discovery is that in SAJO, the interpolation in iteration
6-8 is not monotonic. Sometimes, this non-monotonicity is acceptable
since the target basis does not necessarily have a higher index value
than the current one, while in some cases, the optimisation will be
better off if these intermediate basis can be used as the start of the
next iteration (see section \ref{monotonic} for a detailed discussion on
this).

@@

\caption{Six frames selected from the animation of search paths in the PCA-projected basis space. With animation, the start and end position of each optimisation routine is clearly identified. The full video of the animation can be found at \url{https://vimeo.com/user132007430/review/504242845/b73f37175a}}\label{fig:toy-pca-animated}
\end{figure}
\end{Schunk}

@@

\caption{Six frames selected from the tour view in the original 5D space for the same 1D projection problem as Figure \ref{fig:toy-pca} and \ref{fig:toy-pca-animated}. The full video of the animation can be found at \url{https://vimeo.com/user132007430/review/504328122/9be84db563}}\label{fig:toy-tour}
\end{figure}
\end{Schunk}

>>>>

In projection pursuit the optimisation problem aims at finding the
global and local maxima that give interesting projections according to
an index function. That is, it starts with a given randomly selected
basis \(\mathbf{A}_1\) and aims at finding an optimal final projection
basis \(\mathbf{A}_T\) that satisfies the following optimisation
problem:

@@

\noindent where \(f\), \(\mathcal{A}\), and \(\mathbf{X}\) are defined
as in the previous section, \(I_d\) is the \(d\)-dimensional identity
matrix and the constraint ensures the projection bases \(\mathbf{A}\) to
be orthonormal. It is worth noticing the following: 1) The optimisation
is constrained and the basis space is a \(d\)-dimensional sphere. 2)
There may be index functions for which the objective function might not
be differentiable. 3) While finding the global optimum is the goal of
the optimisation problem, interesting projections may also appear in the
local optimum. 4) The optimisation should be computationally agile since
the tour animation is viewed by the users during the optimisation.

@@

\caption[An inspection of the index values as the optimisation progress for two optimisers]{An inspection of the index values as the optimisation progress for two optimisers: SA (left) and SAJO (right) using trace plot.The holes index is optimised for a 2D projection problem on the six-variable dataset \code{boa6}. The lines indicate the interpolation and the dots indicate new targets generated by the optimisers. The interpolation with both optimisers is smooth while the in the left panel, SAJO is observed to first pass through some bases with higher index value before reach the target bases during Time 76 to 130.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

Another interesting diagnostic is to examine how the index value
improves between interpolating bases since the projection on these bases
is played by the tour animation. Trace plots are created by plotting the
index value against time. Figure \ref{fig:toy-interp} presents the trace
plot of the same optimisers as Figure \ref{fig:toy-search} and one can
observe that the trace is smooth in both cases. It may seem bizarre at
the first sight that the interpolation in the left panel first passes
some bases with higher index value before decreases to a lower target.
This happens because on the one hand, the probabilistic acceptance in
SAJO implies that some worse bases will be accepted by the optimiser. On
the other hand, guided tour interpolate between the current and target
basis to provide a smooth transition between projections. These two
facts indicate that a non-monotonic interpolation can't be avoided for
SAJO. Later, in section \ref{monotonic}, there will be a discussion on
improving the non-monotonic interpolation for SA.

@@

\caption{Six frames selected from the animation of search paths in the PCA-projected basis space. With animation, the start and end position of each optimisation routine is clearly identified. The full video of the animation can be found at \url{https://vimeo.com/user132007430/review/504242845/b73f37175a}.}\label{fig:toy-pca-animated}
\end{figure}
\end{Schunk}

@@

\caption{Six frames selected from the tour view in the original 5D space for the same 1D projection problem as Figure \ref{fig:toy-pca} and \ref{fig:toy-pca-animated}. The full video of the animation can be found at \url{https://vimeo.com/user132007430/review/504328122/9be84db563}.}\label{fig:toy-tour}
\end{figure}
\end{Schunk}
